\thetitle

### 2020-11-21 writing/discussion.tex
Rewording and grammar fixes in discussion

The analysis in the previous section suggests a number of interesting
high-level conclusions about blame in the gradual typing setting.  First,
run time type checks have a large positive impact. This is the case
regardless of whether these checks issue blame or throw plain exceptions.
Second, in many cases type checks that issue blame are more helpful than
those that do not. However, our results indicate that in other cases blame
is not critical and thus it is worth further investigation whether
disabling blame tracking for the sake of performance may be   an
acceptable trade off. Third, from the three different blame strategies the
Natural approach of Typed Racket fares better than the Transient approach
of Reticulated Python but by a small margin. Since Natural offers complete
and sound path-based blame while Transient incomplete but sound heap-based
blame~\cite{gfd-oopsla-2019}, our results call for a deeper understanding
of the relative strengths of the two models for blame.  Fourth, given that
Transient's sound but shallow run time type checks do not seem to hamper
debugging, a version of Typed Racket that disables some of its wrappers
may offer an answer to the performance issues of the Natural semantics. 
Fifth, just as for Typed Racket, the experiment points to possible
improvements for Reticulated Python. In particular, the experimental 
data indicates that the usefulness of transient blame does not depend on
whether  the rational programmer follows first or the last element from a
Transient blame set. Hence, given the performance issues of Transient we discuss further on, 
a possible optimization is to limit the size of blame sets
using timestamps.

There are a number of threats to the validity of our conclusions that we
describe in the previous section our attempts to mitigate: (i) the
representativeness of benchmarks; (ii) the relation between mutations and
real programming mistakes; (iii) the definition of interesting debugging
scenarios; and (iv) our approach to sampling scenarios. In the remainder
of this section, we discuss extensively two additional threats.

@@

As we discuss in section~\ref{sec:results}, surprisingly Transient blames
an empty set of boundaries in 967 scenarios.
Such cases should never occur, at least in theory, because an empty set
 means the value has never crossed a boundary---if the value is indeed defined
 in the current module, then we have typed code blaming itself for a typed
 value as all the checks are inlined in typed code. Thus, the type checker must be unsound.
After careful investigation of these empty blame cases, we found no soundness
 bugs in Transient Typed Racket.
Instead, we found scenarios in which Transient lost track of the proper
 boundaries.
These scenarios suggest ways to improve~\citet{vss-popl-2017}'s blame
algorithm:
\begin{itemize}
  \item
    Entries in a blame map must point to several parent entries.
    For example, if the function \texttt{f} receives bad input in the call
    \texttt{(filter f xs)}, then blame should point to both \texttt{filter}
    and the \texttt{xs} list.
  \item
    The construction of blame map entries must be guided by type-like specifications
     instead of relying on syntax.
    Aliasing the built-in \texttt{filter} function should not change the shape
     of the blame map.
  \item
    The initial blame map must reflect the initial type environment.
    For instance, Typed Racket trusts that untyped core-library functions, such as \texttt{filter},
     behave correctly; such assumptions should appear in the blame map.
\end{itemize}
\noindent{}Despite these known limitations, based on the relatively small
number of debugging failures, we conjecture that these improvements
 to Transient blame-tracking do not affect our overall conclusions.

A second apparent flaw in our Transient semantics is the high cost of blame.
Whereas \citet{vss-popl-2017} report an average slowdown of 2.5x and
worst-case of 5.4x on fully-typed benchmarks in Reticulated,
some of our configurations exceed a 4 minute timeout or 6GB memory limit with transient blame.
To put those limits into context, the fully typed and fully untyped benchmarks all complete in a few seconds with minimal memory usage, and none of the mixed Natural configurations hit these limits.
Unfortunately, our high cost is closer to the truth;
 there are at least three broad issues that skew the earlier results.
First, the 2017 implementation of Reticulated fails to insert certain
 soundness checks\footnote{Missing check: \url{https://github.com/mvitousek/reticulated/issues/36}}
 and blame-map updates\footnote{Missing cast: \url{https://github.com/mvitousek/reticulated/issues/43}}
 from the paper.
Second, Reticulated's type checker assigns the dynamic type to local
 variables in the "fully typed" benchmarks---often because the type system
 cannot articulate a more precise type.
Code that ends up with the dynamic type has fewer constraints to check at run-time.
Third, \citet{vss-popl-2017} use relatively small benchmarks.
Four have since been retired from the official Python benchmark suite
 because they are too small, unrealistic, and unstable.\footnote{Release notes: \url{https://pyperformance.readthedocs.io/changelog.html}}
On larger programs, Reticulated suffers from high overhead.
For example, a Reticulated version of our \texttt{sieve} benchmark runs in
 ~40 seconds without blame, and with blame times out after 10 minutes.

>>>>

The analysis in the previous section suggests a number of interesting
high-level conclusions about blame in the gradual typing setting.  First,
run time type checks have a large positive impact. This is the case
regardless of whether these checks issue blame or throw plain exceptions.
Second, in many cases type checks that issue blame are more helpful than
those that do not. However, our results indicate that in other cases blame
is not critical and thus it is worth further investigation whether
disabling blame tracking for the sake of performance may be   an
acceptable trade off. Third, from the three different blame strategies, the
Natural approach of Typed Racket fares better than the Transient approach
of Reticulated Python, but only by a small margin. Since Natural offers complete
and sound path-based blame while Transient offers incomplete but sound heap-based
blame~\cite{gfd-oopsla-2019}, our results call for a deeper understanding
of the relative strengths of the two models for blame.  Fourth, given that
Transient's sound but shallow run time type checks do not seem to hamper
debugging, a version of Typed Racket that disables some of its wrappers
may offer an answer to the performance issues of the Natural semantics. 
Fifth, just as for Typed Racket, the experiment points to possible
improvements for Reticulated Python. In particular, the experimental 
data indicates that the usefulness of transient blame does not depend on
whether  the rational programmer follows first or the last element from a
Transient blame set. Hence, given the performance issues of Transient we discuss further on, 
a possible optimization is to limit the size of blame sets
using timestamps.

There are a number of threats to the validity of our conclusions: (i) the
representativeness of benchmarks; (ii) the relation between mutations and
real programming mistakes; (iii) the definition of interesting debugging
scenarios; and (iv) our approach to sampling scenarios. The previous sections
have described the various ways we attempt to mitigate these threats;
we discuss two additional threats in the remainder of this section.

@@

As we discuss in section~\ref{sec:results}, Transient surprisingly blames
an empty set of boundaries in 967 scenarios.
Such cases should never occur, at least in theory, because an empty set
 means the value has never crossed a boundary---if the value is indeed defined
 in the current module, then we have typed code blaming itself for a typed
 value as all the checks are inlined in typed code. Thus, the type checker must be unsound.
After careful investigation of these empty blame cases, we found no soundness
 bugs in Transient Typed Racket.
Instead, we found scenarios in which Transient lost track of the proper
 boundaries.
These scenarios suggest ways to improve~\citet{vss-popl-2017}'s blame
algorithm:
\begin{itemize}
  \item
    Entries in a blame map must point to several parent entries.
    For example, if the function \texttt{f} receives bad input in the call
    \texttt{(filter f xs)}, then blame should point to both \texttt{filter}
    and the \texttt{xs} list.
  \item
    The construction of blame map entries must be guided by type-like specifications
     instead of relying on syntax.
    Aliasing the built-in \texttt{filter} function should not change the shape
     of the blame map.
  \item
    The initial blame map must reflect the initial type environment.
    For instance, Typed Racket trusts that untyped core-library functions, such as \texttt{filter},
     behave correctly; such assumptions should appear in the blame map.
\end{itemize}
\noindent{}Despite these known limitations, based on the relatively small
number of debugging failures, we conjecture that these improvements
 to Transient blame-tracking do not affect our overall conclusions.

A second apparent flaw in our Transient semantics is the high cost of blame.
Whereas \citet{vss-popl-2017} report an average slowdown of 2.5x and
worst-case of 5.4x on fully-typed benchmarks in Reticulated,
some of our configurations exceed our 4 minute timeout or 6GB memory limit with transient blame.
To put those limits into context, the fully typed and fully untyped benchmarks all complete in a few seconds with minimal memory usage, and none of the mixed Natural configurations hit these limits.
Unfortunately, our high cost for Transient seems closer to the truth;
 there are at least three broad issues that skew the earlier results.
First, the 2017 implementation of Reticulated fails to insert certain
 soundness checks\footnote{Missing check: \url{https://github.com/mvitousek/reticulated/issues/36}}
 and blame-map updates\footnote{Missing cast: \url{https://github.com/mvitousek/reticulated/issues/43}}
 from the paper.
Second, Reticulated's type checker assigns the dynamic type to local
 variables in the "fully typed" benchmarks---often because the type system
 cannot articulate a more precise type.
Code that ends up with the dynamic type has fewer constraints to check at run-time.
Third, \citet{vss-popl-2017} use relatively small benchmarks.
Four have since been retired from the official Python benchmark suite
 because they are too small, unrealistic, and unstable.\footnote{Release notes: \url{https://pyperformance.readthedocs.io/changelog.html}}
On larger programs, Reticulated suffers from high overhead.
For example, a Reticulated version of our simplest benchmark runs in
 ~40 seconds without blame, and with blame times out after 10 minutes.
